---
title: "TreeSE_Coxmodel.qmd"
format: html
editor: visual
---

## This report focuses on evaluating the predictive performance of the Cox proportional hazards model under transformed and un-transformed data. The primary objective is to identify which transformation method allows the Cox model to yield the most accurate survival predictions. By systematically applying distinct transformations to the data-set and assessing the model’s performance, this study aims to provide insights into optimizing data pre-processing for survival analysis.

Load libraries

```{r, message=FALSE, warning=FALSE, echo=TRUE}
library(mia)
library(ggplot2)
library(lattice)
library(TreeSummarizedExperiment)
library(coda4microbiome)
library(SummarizedExperiment)
library(tibble)
library(phyloseq)
library(survival)
library(survival)
```

link to download the biom.Rdata dataset: https://github.com/wg99526/MiSurvGit/tree/main/Data

```{r}
load("~/Desktop/Microbiome_Analysis/TreeSE/biom.Rdata")
```

convert phyloseq object to TreeSE

```{r, message=FALSE, warning=FALSE, echo=TRUE}
tse <- convertFromPhyloseq(biom)
```

Transform the data

```{r}
coef_threshold = 0
lambda = "lambda.1se" 
nvar = NULL
alpha = 0.9
nfolds = 10
showPlots = TRUE
x1 <- impute_zeros(assay(tse))
x1 <- t(x1)
kselect <- ncol(x1)
taxaselect <- (1:ncol(x1))
lrmatrix <- logratios_matrix(x1)
lrX <- lrmatrix[[1]]
idlrX <- lrmatrix[[2]]
```

split the data into training and testing subsets

```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Set seed for reproducibility
set.seed(123)
# Split TreeSE into training and test sets by samples (columns)
num_samples <- ncol(tse)
train_indices <- sample(seq_len(num_samples), size = floor(0.7 * num_samples))
test_indices <- setdiff(seq_len(num_samples), train_indices)
# Subset TreeSE
tse_train <- tse[, train_indices]
tse_test <- tse[, test_indices]
```

Prepare training data

```{r}
# Extract colData
col_data_train <- as.data.frame(colData(tse_train))
x <- t(assay(tse_train))
# Extract time-to-event and status from colData
time <- col_data_train$T1Dweek   # numeric vector
status <- col_data_train$T1D  # binary vector (0 = censored, 1 = event)
covars <- col_data_train[, c("Antibiotics", "Sex")]  # update with correct column names
# Assuming lrX is a matrix or data frame with columns
num_rows <- nrow(lrX)
split_point <- floor(0.7 * num_rows)

lrX_train <- lrX[1:split_point, ]



# Replace row names with only digits extracted from original row names
rownames(x) <- gsub("\\D", "", rownames(x))
colnames(x) <- gsub("\\D", "", colnames(x))
```

Prepare testing data

```{r}
#prepare the testing subset
col_data_test <- as.data.frame(colData(tse_test))
# Extract time-to-event and status from colData
time_test <- col_data_test$T1Dweek   # numeric vector
status_test <- col_data_test$T1D  # binary vector (0 = censored, 1 = event)
covars_test <- col_data_test[, c("Antibiotics", "Sex")]  # update with correct column names
# Extract abundance matrix (assumes 1st assay)
x_test <- t(assay(tse_test))  # transpose: rows = samples, columns = features
lrX_test <- lrX[(split_point + 1):num_rows, ]
# Replace row names with only digits extracted from original row names
rownames(x_test) <- gsub("\\D", "", rownames(x_test))
colnames(x_test) <- gsub("\\D", "", colnames(x_test))
```

implement cox model

```{r}
set.seed(123)
y_test = Surv(time_test, status_test)# Cox response variable (time & status)
y = Surv(time, status)# fits a standard Cox proportional hazards model (coxph) using all covariates
df0 <- data.frame(as.matrix(y), covars)
model0 <- coxph(Surv(time, status) ~ ., data = df0)
x0 <- predict(model0)#linear predictor
#Penalized Cox Model with Offset
cvfit <- glmnet::cv.glmnet(lrX_train, y, family = "cox",
                           type.measure = "C",
                           nfolds = nfolds, alpha = alpha,
                           keep = TRUE,
                           offset = x0)
plot(cvfit)
predictions<-as.numeric(predict(cvfit, lrX_test, s = lambda, newoffset=x0))
Cindex_signature <- glmnet::Cindex(pred = predictions, y_test)
```

Generate plot for transformed data

```{r}
rowlasso <- max(which(cvfit$glmnet.fit$df <= nvar))
lambdavalue <- lambda
  if (is.character(lambda)) {
    if (lambda == "lambda.1se")
      lambdavalue <- cvfit$lambda.1se
    if (lambda == "lambda.min")
      lambdavalue <- cvfit$lambda.min
  }
idrow <- max(which(cvfit$glmnet.fit$lambda >= lambdavalue))
coeflr <- as.vector(coef(cvfit, s = lambda)) #[-1]
lrselect <- which(coeflr != 0)
coeflogcontrast <- rep(0, ncol(x))
for (i in (1:length(coeflr))) {
coeflogcontrast[idlrX[i, 1]] <- coeflogcontrast[idlrX[i,1]] +  coeflr[i]
coeflogcontrast[idlrX[i, 2]] <- coeflogcontrast[idlrX[i,2]] -   coeflr[i]
  }
varlogcontrast <- which(abs(coeflogcontrast) > coef_threshold)
coeflogcontrast <- coeflogcontrast[varlogcontrast]

(names.select <- colnames(x)[varlogcontrast])
(sign <- ifelse(coeflogcontrast > 0, 1, 0))
sign <- factor(sign, levels = c(0, 1), labels = c("negative",
                                                    "positive"))


#predictions<-x0+logcontrast
coeflogcontrast<-2*coeflogcontrast/sum(abs(coeflogcontrast))


mcvCindex <- cvfit$cvm[idrow]
sdcvCindex <- cvfit$cvsd[idrow]

plot1<-NULL
plot2<-NULL

if (length(lrselect>0)){

plot1 <- plot_riskscore(predictions, x_test, time_test, status_test, showPlots = showPlots)
plot2 <- plot_signature(names.select, coeflogcontrast, showPlots = showPlots)

  } else {
    print("No variables are selected. The risk score plot and the signature plot are not displayed.")
  }
```

implement cox model on un-transformed

```{r}
set.seed(123)

# fits a standard Cox proportional hazards model (coxph) using all covariates
df0 <- data.frame(as.matrix(y), covars)
model1 <- coxph(Surv(time, status) ~ ., data = df0)
x0 <- predict(model1)#linear predictor
#Penalized Cox Model with Offset
cvfit1 <- glmnet::cv.glmnet(x, y, family = "cox",
                            type.measure = "C",
                            nfolds = nfolds, alpha = alpha,
                            keep = TRUE,
                            offset = x0)
plot(cvfit1)
predictions1<-as.numeric(predict(cvfit1,x_test,s=lambda, newoffset=x0))
Cindex_signature <- glmnet::Cindex(pred = predictions1, y_test)
```

Generate plots for untransformed

```{r}
rowlasso <- max(which(cvfit$glmnet.fit$df <= nvar))

lambdavalue <- lambda
if (is.character(lambda)) {
  if (lambda == "lambda.1se")
    lambdavalue <- cvfit$lambda.1se
  if (lambda == "lambda.min")
    lambdavalue <- cvfit$lambda.min
}

idrow <- max(which(cvfit$glmnet.fit$lambda >= lambdavalue))
coeflr <- as.vector(coef(cvfit, s = lambda))
lrselect <- which(coeflr != 0)

coeflogcontrast <- rep(0, ncol(x))
for (i in (1:length(coeflr))) {
  coeflogcontrast[idlrX[i, 1]] <- coeflogcontrast[idlrX[i, 1]] + coeflr[i]
  coeflogcontrast[idlrX[i, 2]] <- coeflogcontrast[idlrX[i, 2]] - coeflr[i]
}

varlogcontrast <- which(abs(coeflogcontrast) > coef_threshold)
coeflogcontrast <- coeflogcontrast[varlogcontrast]

names.select <- colnames(x)[varlogcontrast]
sign <- ifelse(coeflogcontrast > 0, 1, 0)
sign <- factor(sign, levels = c(0, 1), labels = c("negative", "positive"))


mcvCindex <- cvfit1$cvm[idrow]
sdcvCindex <- cvfit$cvsd[idrow]

plot1 <- NULL
plot2 <- NULL

if (length(lrselect) > 0) {
  plot1 <- plot_riskscore(predictions, x_test, time_test, status_test, showPlots = showPlots)
  plot2 <- plot_signature(names.select, coeflogcontrast, showPlots = showPlots)
} else {
  print("No variables are selected. The risk score plot and the signature plot are not displayed.")
}

```

Evaluate model performance on transformed and un-transformed data

```{r}
# Set seed for reproducibility
set.seed(123)

# Define the models and their corresponding test matrices
models <- list(
 transformed = list(model = model0, cvfit = cvfit, x_test = lrX_test, x_train = lrX),
 untransformed = list(model = model1, cvfit = cvfit1, x_test = x_test, x_train = x)
)

# Prepare result containers
cindex_results <- list()          # Stores the point estimate of the C-index for each model
cindex_bootstrap <- list()        # Stores the bootstrap-based statistics for each model

# Set number of bootstrap replicates
n_boot <- 1000

# Loop through each model type: 'transformed' and 'untransformed'
for (type in names(models)) {
  cat("\n--- Testing:", type, "---\n")

  # Extract the model and data for the current type
  current_model <- models[[type]]$model
  current_cvfit <- models[[type]]$cvfit
  current_x_test <- models[[type]]$x_test

  # Identify complete cases (no missing values in covariates, test data, or survival outcome)
  complete_cases <- complete.cases(covars_test) &
                    complete.cases(current_x_test) &
                    !is.na(time_test) & !is.na(status_test)

  # Subset testing data based on complete cases
  covars_test_clean <- covars_test[complete_cases, , drop = FALSE]
  x_test_clean <- current_x_test[complete_cases, , drop = FALSE]
  time_test_clean <- time_test[complete_cases]
  status_test_clean <- status_test[complete_cases]

  # Predict linear predictors using covariates-only model (for use as offsets)
  x0_test_clean <- predict(current_model,
                           newdata = as.data.frame(covars_test_clean),
                           type = "lp")

  # Predict risk scores using penalized Cox model with offset
  lambda_value <- current_cvfit$lambda.min
  test_predictions <- predict(current_cvfit,
                              newx = x_test_clean,
                              s = lambda_value,
                              newoffset = x0_test_clean,
                              type = "link")

  # Flip sign for correct direction of risk (higher score = higher risk)
  test_predictions <- -as.numeric(test_predictions)

  # Compute point estimate of Harrell's C-index
  test_cindex <- glmnet::Cindex(pred = test_predictions,
                                y = Surv(time_test_clean, status_test_clean))
  cindex_results[[type]] <- test_cindex

  # Perform bootstrap to estimate confidence interval of C-index
  boot_cindex <- numeric(n_boot)
  for (b in 1:n_boot) {
    idx <- sample(seq_along(time_test_clean), replace = TRUE)  # sample with replacement
    boot_cindex[b] <- glmnet::Cindex(pred = test_predictions[idx],
                                     y = Surv(time_test_clean[idx], status_test_clean[idx]))
  }

  # Store bootstrap statistics
  cindex_bootstrap[[type]] <- list(
    mean = mean(boot_cindex),
    ci = quantile(boot_cindex, probs = c(0.025, 0.975)),  # 95% confidence interval
    all = boot_cindex  # full vector of bootstrapped C-index values
  )

  # Print results
  cat("Samples used:", length(time_test_clean), "\n")
  cat("C-index:", round(test_cindex, 4), "\n")
  cat("Bootstrap 95% CI:", round(cindex_bootstrap[[type]]$ci[1], 4), "-", round(cindex_bootstrap[[type]]$ci[2], 4), "\n")
}

```
