---
title: "TreeSE_Coxmodel.qmd"
format: html
editor: visual
---

## This report focuses on evaluating the predictive performance of the Cox proportional hazards model under transformed and un-transformed data. The primary objective is to identify which transformation method allows the Cox model to yield the most accurate survival predictions. By systematically applying distinct transformations to the data-set and assessing the model’s performance, this study aims to provide insights into optimizing data pre-processing for survival analysis.

Load libraries

```{r, message=FALSE, warning=FALSE, echo=TRUE}
library(mia)
library(ggplot2)
library(lattice)
library(TreeSummarizedExperiment)
library(coda4microbiome)
library(SummarizedExperiment)
library(tibble)
library(phyloseq)
library(survival)
library(survival)
```

link to download the biom.Rdata dataset: https://github.com/wg99526/MiSurvGit/tree/main/Data

```{r}
load("~/Desktop/Microbiome_Analysis/TreeSE/biom.Rdata")
```

convert phyloseq object to TreeSE

```{r, message=FALSE, warning=FALSE, echo=TRUE}
tse <- convertFromPhyloseq(biom)
```

## I-Transform the data to all pairwise log ratio

```{r}
coef_threshold = 0
lambda = "lambda.1se" 
nvar = NULL
alpha = 0.9
nfolds = 10
showPlots = TRUE
x1 <- impute_zeros(assay(tse))
x1 <- t(x1)
kselect <- ncol(x1)
taxaselect <- (1:ncol(x1))
lrmatrix <- logratios_matrix(x1)
lrX <- lrmatrix[[1]]
idlrX <- lrmatrix[[2]]
```

## II- Transform the data to clr data format

```{r, message=FALSE, warning=FALSE, echo=TRUE}
tse_clr <- transformAssay(
    x = tse,
    assay.type = "counts",
    method = "clr",
    pseudocount = TRUE,
    name = "clr"
)
```

## III- Transform the data to relative abundance data format

```{r}
tse_rel <- transformAssay(tse, assay.type = "counts", method = "relabundance")
```

split the data into training and testing subsets

```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Set seed for reproducibility
set.seed(123)
# Split TreeSE into training and test sets by samples (columns)
num_samples <- ncol(tse)
train_indices <- sample(seq_len(num_samples), size = floor(0.7 * num_samples))
test_indices <- setdiff(seq_len(num_samples), train_indices)
# Subset TreeSE
tse_train <- tse[, train_indices]
tse_test <- tse[, test_indices]
#split for clr data
# Set seed for reproducibility
set.seed(123)
# Split TreeSE into training and test sets by samples (columns)
num_samples_clr <- ncol(tse_clr)
train_indices_clr <- sample(seq_len(num_samples_clr), size = floor(0.7 * num_samples_clr))
test_indices_clr <- setdiff(seq_len(num_samples_clr), train_indices_clr)
# Subset TreeSE
tse_clr_train <- tse_clr[, train_indices_clr]
tse_clr_test <- tse_clr[, test_indices_clr]
#split for relative abundance data
set.seed(123)
# Split TreeSE into training and test sets by samples (columns)
num_samples_rel <- ncol(tse_rel)
train_indices_rel <- sample(seq_len(num_samples_rel), size = floor(0.7 * num_samples_rel))
test_indices_rel <- setdiff(seq_len(num_samples_rel), train_indices_rel)
# Subset TreeSE
tse_rel_train <- tse_rel[, train_indices_rel]
tse_rel_test <- tse_rel[, test_indices_rel]
```

Prepare training data

```{r}
# Extract colData
col_data_train <- as.data.frame(colData(tse_train))
x <- t(assay(tse_train))
# Extract time-to-event and status from colData
time <- col_data_train$T1Dweek   # numeric vector
status <- col_data_train$T1D  # binary vector (0 = censored, 1 = event)
covars <- col_data_train[, c("Antibiotics", "Sex")]  # update with correct column names
# Assuming lrX is a matrix or data frame with columns
num_rows <- nrow(lrX)
split_point <- floor(0.7 * num_rows)
lrX_train <- lrX[1:split_point, ]
# Replace row names with only digits extracted from original row names
rownames(lrX_train) <- gsub("\\D", "", rownames(lrX_train))
colnames(lrX_train) <- gsub("\\D", "", colnames(lrX_train))
#training data clr
# Extract colData
col_data_train_clr <- as.data.frame(colData(tse_clr_train))
x_clr <- t(assay(tse_clr_train))
# Extract time-to-event and status from colData
time_clr <- col_data_train_clr$T1Dweek   # numeric vector
status_clr <- col_data_train_clr$T1D  # binary vector (0 = censored, 1 = event)
covars_clr <- col_data_train_clr[, c("Antibiotics", "Sex")]  # update with correct column names
# Replace row names with only digits extracted from original row names
rownames(x_clr) <- gsub("\\D", "", rownames(x_clr))
colnames(x_clr) <- gsub("\\D", "", colnames(x_clr))
#training data rel
# Extract colData
col_data_train_rel <- as.data.frame(colData(tse_rel_train))
x_rel <- t(assay(tse_rel_train))
# Extract time-to-event and status from colData
time_rel <- col_data_train_rel$T1Dweek   # numeric vector
status_rel <- col_data_train_rel$T1D  # binary vector (0 = censored, 1 = event)
covars_rel <- col_data_train_rel[, c("Antibiotics", "Sex")]  # update with correct column names
# Replace row names with only digits extracted from original row names
rownames(x_rel) <- gsub("\\D", "", rownames(x_rel))
colnames(x_rel) <- gsub("\\D", "", colnames(x_rel))
```

Prepare testing data

```{r}
#prepare the testing subset
col_data_test <- as.data.frame(colData(tse_test))
# Extract time-to-event and status from colData
time_test <- col_data_test$T1Dweek   # numeric vector
status_test <- col_data_test$T1D  # binary vector (0 = censored, 1 = event)
covars_test <- col_data_test[, c("Antibiotics", "Sex")]  # update with correct column names
# Extract abundance matrix (assumes 1st assay)
x_test <- t(assay(tse_test))  # transpose: rows = samples, columns = features
lrX_test <- lrX[(split_point + 1):num_rows, ]
# Replace row names with only digits extracted from original row names
rownames(x_test) <- gsub("\\D", "", rownames(x_test))
colnames(x_test) <- gsub("\\D", "", colnames(x_test))
#testing data for clr
col_data_test_clr <- as.data.frame(colData(tse_clr_test))
# Extract time-to-event and status from colData
time_test_clr <- col_data_test_clr$T1Dweek   # numeric vector
status_test_clr <- col_data_test_clr$T1D  # binary vector (0 = censored, 1 = event)
covars_test_clr <- col_data_test_clr[, c("Antibiotics", "Sex")]  # update with correct column names
# Extract abundance matrix (assumes 1st assay)
x_test_clr <- t(assay(tse_clr_test))  # transpose: rows = samples, columns = features
lrX_test <- lrX[(split_point + 1):num_rows, ]
# Replace row names with only digits extracted from original row names
rownames(x_test_clr) <- gsub("\\D", "", rownames(x_test_clr))
colnames(x_test_clr) <- gsub("\\D", "", colnames(x_test_clr))
#testing data rel
col_data_test_rel <- as.data.frame(colData(tse_rel_test))
# Extract time-to-event and status from colData
time_test_rel <- col_data_test_rel$T1Dweek   # numeric vector
status_test_rel <- col_data_test_rel$T1D  # binary vector (0 = censored, 1 = event)
covars_test_rel <- col_data_test_rel[, c("Antibiotics", "Sex")]  # update with correct column names
# Extract abundance matrix (assumes 1st assay)
x_test_rel <- t(assay(tse_rel_test))  # transpose: rows = samples, columns = features
lrX_test <- lrX[(split_point + 1):num_rows, ]
# Replace row names with only digits extracted from original row names
rownames(x_test_rel) <- gsub("\\D", "", rownames(x_test_rel))
colnames(x_test_rel) <- gsub("\\D", "", colnames(x_test_rel))
```

implement cox model on: all pair-wise log ratio transformation/ clr transformation / raw data / relative abundance

```{r}
#1-all pair-wise log ratio
set.seed(123)
y_test = Surv(time_test, status_test)# Cox response variable (time & status)
y = Surv(time, status)# fits a standard Cox proportional hazards model (coxph) using all covariates
df1 <- data.frame(as.matrix(y), covars)
model1 <- coxph(Surv(time, status) ~ ., data = df1)
x1 <- predict(model1)#linear predictor
#Penalized Cox Model with Offset
cvfit1 <- glmnet::cv.glmnet(lrX_train, y, family = "cox",
                           type.measure = "C",
                           nfolds = nfolds, alpha = alpha,
                           keep = TRUE,
                           offset = x1)
plot(cvfit1, main = "All pairwise-log ratio", line = 3)
predictions1<-as.numeric(predict(cvfit1, lrX_test, s = lambda, newoffset=x1))
Cindex_signature1 <- glmnet::Cindex(pred = predictions1, y_test)
#2-raw data
set.seed(123)
# fits a standard Cox proportional hazards model (coxph) using all covariates
df2 <- data.frame(as.matrix(y), covars)
model2 <- coxph(Surv(time, status) ~ ., data = df2)
x2 <- predict(model2)#linear predictor
#Penalized Cox Model with Offset
cvfit2 <- glmnet::cv.glmnet(x, y, family = "cox",
                            type.measure = "C",
                            nfolds = nfolds, alpha = alpha,
                            keep = TRUE,
                            offset = x2)
plot(cvfit2, main = "original data", line = 3)
predictions2<-as.numeric(predict(cvfit2,x_test,s=lambda, newoffset=x1))
Cindex_signature2 <- glmnet::Cindex(pred = predictions2, y_test)
#3-clr
set.seed(123)
y_test_clr = Surv(time_test_clr, status_test_clr)# Cox response variable (time & status)
y_clr = Surv(time_clr, status_clr)# fits a standard Cox proportional hazards model (coxph) using all covariates
df3 <- data.frame(as.matrix(y_clr), covars_clr)
model3 <- coxph(Surv(time_clr, status_clr) ~ ., data = df3)
x3 <- predict(model3)#linear predictor
#Penalized Cox Model with Offset
cvfit3 <- glmnet::cv.glmnet(x_clr, y_clr, family = "cox",
                           type.measure = "C",
                           nfolds = nfolds, alpha = alpha,
                           keep = TRUE,
                           offset = x3)
plot(cvfit3, main = "clr",line = 3)
predictions3<-as.numeric(predict(cvfit3, x_test_clr, s = lambda, newoffset=x3))
Cindex_signature3 <- glmnet::Cindex(pred = predictions3, y_test_clr)
#4-relative abundance
set.seed(123)
y_test_rel = Surv(time_test_rel, status_test_rel)# Cox response variable (time & status)
y_rel = Surv(time_rel, status_rel)# fits a standard Cox proportional hazards model (coxph) using all covariates
df4 <- data.frame(as.matrix(y_rel), covars_rel)
model4 <- coxph(Surv(time_rel, status_rel) ~ ., data = df4)
x4 <- predict(model4)#linear predictor
#Penalized Cox Model with Offset
cvfit4 <- glmnet::cv.glmnet(x_rel, y_rel, family = "cox",
                           type.measure = "C",
                           nfolds = nfolds, alpha = alpha,
                           keep = TRUE,
                           offset = x4)
plot(cvfit4, main = "relative abundance", line = 3)
predictions4<-as.numeric(predict(cvfit4, x_test_rel, s = lambda, newoffset=x4))
Cindex_signature4 <- glmnet::Cindex(pred = predictions4, y_test_rel)
```

Generate plot for all pair-wise log ratio transformation

```{r}
rowlasso <- max(which(cvfit1$glmnet.fit$df <= nvar))
lambdavalue <- lambda
  if (is.character(lambda)) {
    if (lambda == "lambda.1se")
      lambdavalue <- cvfit1$lambda.1se
    if (lambda == "lambda.min")
      lambdavalue <- cvfit1$lambda.min
  }
idrow <- max(which(cvfit1$glmnet.fit$lambda >= lambdavalue))
coeflr <- as.vector(coef(cvfit1, s = lambda)) #[-1]
lrselect <- which(coeflr != 0)
coeflogcontrast <- rep(0, ncol(x))
for (i in (1:length(coeflr))) {
coeflogcontrast[idlrX[i, 1]] <- coeflogcontrast[idlrX[i,1]] +  coeflr[i]
coeflogcontrast[idlrX[i, 2]] <- coeflogcontrast[idlrX[i,2]] -   coeflr[i]
  }
varlogcontrast <- which(abs(coeflogcontrast) > coef_threshold)
coeflogcontrast <- coeflogcontrast[varlogcontrast]

(names.select <- colnames(x)[varlogcontrast])
(sign <- ifelse(coeflogcontrast > 0, 1, 0))
sign <- factor(sign, levels = c(0, 1), labels = c("negative",
                                                    "positive"))


#predictions<-x0+logcontrast
coeflogcontrast<-2*coeflogcontrast/sum(abs(coeflogcontrast))


mcvCindex <- cvfit1$cvm[idrow]
sdcvCindex <- cvfit1$cvsd[idrow]

plot1<-NULL
plot2<-NULL

if (length(lrselect>0)){

plot1 <- plot_riskscore(predictions1, x_test, time_test, status_test, showPlots = showPlots)
plot2 <- plot_signature(names.select, coeflogcontrast, showPlots = showPlots)

  } else {
    print("No variables are selected. The risk score plot and the signature plot are not displayed.")
  }
```

Generate plots for original data

```{r}
rowlasso <- max(which(cvfit2$glmnet.fit$df2 <= nvar))

lambdavalue <- lambda
if (is.character(lambda)) {
  if (lambda == "lambda.1se")
    lambdavalue <- cvfit2$lambda.1se
  if (lambda == "lambda.min")
    lambdavalue <- cvfit2$lambda.min
}

idrow <- max(which(cvfit2$glmnet.fit$lambda >= lambdavalue))
coeflr <- as.vector(coef(cvfit2, s = lambda))
lrselect <- which(coeflr != 0)

coeflogcontrast <- rep(0, ncol(x))
for (i in (1:length(coeflr))) {
  coeflogcontrast[idlrX[i, 1]] <- coeflogcontrast[idlrX[i, 1]] + coeflr[i]
  coeflogcontrast[idlrX[i, 2]] <- coeflogcontrast[idlrX[i, 2]] - coeflr[i]
}

varlogcontrast <- which(abs(coeflogcontrast) > coef_threshold)
coeflogcontrast <- coeflogcontrast[varlogcontrast]

names.select <- colnames(x)[varlogcontrast]
sign <- ifelse(coeflogcontrast > 0, 1, 0)
sign <- factor(sign, levels = c(0, 1), labels = c("negative", "positive"))


mcvCindex <- cvfit2$cvm[idrow]
sdcvCindex <- cvfit2$cvsd[idrow]

plot1 <- NULL
plot2 <- NULL

if (length(lrselect) > 0) {
  plot1 <- plot_riskscore(predictions2, x_test, time_test, status_test, showPlots = showPlots)
  plot2 <- plot_signature(names.select, coeflogcontrast, showPlots = showPlots)
} else {
  print("No variables are selected. The risk score plot and the signature plot are not displayed.")
}
```

generate plots for clr transformation

```{r}
rowlasso <- max(which(cvfit3$glmnet.fit$df3 <= nvar))

lambdavalue <- lambda
if (is.character(lambda)) {
  if (lambda == "lambda.1se")
    lambdavalue <- cvfit3$lambda.1se
  if (lambda == "lambda.min")
    lambdavalue <- cvfit3$lambda.min
}

idrow <- max(which(cvfit3$glmnet.fit$lambda >= lambdavalue))
coeflr <- as.vector(coef(cvfit3, s = lambda))
lrselect <- which(coeflr != 0)

coeflogcontrast <- rep(0, ncol(x))
for (i in (1:length(coeflr))) {
  coeflogcontrast[idlrX[i, 1]] <- coeflogcontrast[idlrX[i, 1]] + coeflr[i]
  coeflogcontrast[idlrX[i, 2]] <- coeflogcontrast[idlrX[i, 2]] - coeflr[i]
}

varlogcontrast <- which(abs(coeflogcontrast) > coef_threshold)
coeflogcontrast <- coeflogcontrast[varlogcontrast]

names.select <- colnames(x)[varlogcontrast]
sign <- ifelse(coeflogcontrast > 0, 1, 0)
sign <- factor(sign, levels = c(0, 1), labels = c("negative", "positive"))


mcvCindex <- cvfit3$cvm[idrow]
sdcvCindex <- cvfit3$cvsd[idrow]

plot1 <- NULL
plot2 <- NULL

if (length(lrselect) > 0) {
  plot1 <- plot_riskscore(predictions2, x_test, time_test, status_test, showPlots = showPlots)
  plot2 <- plot_signature(names.select, coeflogcontrast, showPlots = showPlots)
} else {
  print("No variables are selected. The risk score plot and the signature plot are not displayed.")
}
```

generate plots for relative abundance

```{r}
lambdavalue <- lambda
if (is.character(lambda)) {
  if (lambda == "lambda.1se")
    lambdavalue <- cvfit4$lambda.1se
  if (lambda == "lambda.min")
    lambdavalue <- cvfit4$lambda.min
}

idrow <- max(which(cvfit3$glmnet.fit$lambda >= lambdavalue))
coeflr <- as.vector(coef(cvfit4, s = lambda))
lrselect <- which(coeflr != 0)

coeflogcontrast <- rep(0, ncol(x))
for (i in (1:length(coeflr))) {
  coeflogcontrast[idlrX[i, 1]] <- coeflogcontrast[idlrX[i, 1]] + coeflr[i]
  coeflogcontrast[idlrX[i, 2]] <- coeflogcontrast[idlrX[i, 2]] - coeflr[i]
}

varlogcontrast <- which(abs(coeflogcontrast) > coef_threshold)
coeflogcontrast <- coeflogcontrast[varlogcontrast]

names.select <- colnames(x)[varlogcontrast]
sign <- ifelse(coeflogcontrast > 0, 1, 0)
sign <- factor(sign, levels = c(0, 1), labels = c("negative", "positive"))


mcvCindex <- cvfit4$cvm[idrow]
sdcvCindex <- cvfit4$cvsd[idrow]

plot1 <- NULL
plot2 <- NULL

if (length(lrselect) > 0) {
  plot1 <- plot_riskscore(predictions2, x_test_clr, time_test, status_test, showPlots = showPlots)
  plot2 <- plot_signature(names.select, coeflogcontrast, showPlots = showPlots)
} else {
  print("No variables are selected. The risk score plot and the signature plot are not displayed.")
}
```

```{r}
# Collect C-index results in a named list
cindex_results <- list(
  all_pairwise_log_ratio = Cindex_signature1,
  raw_data               = Cindex_signature2,
  clr_transformation     = Cindex_signature3,
  relative_abundance     = Cindex_signature4
)

# Print C-index values
for (name in names(cindex_results)) {
  cat(paste0(name, ": Harrell's C-index = ", round(cindex_results[[name]], 4), "\n"))
}
```
